### 2020-01-23 interim/sections/background/codegen.tex

\subsection{Type-Level Encoding}
\begin{itemize}
\item PureScript implementation that addresses the 2 criteria statically
\begin{itemize}
\item Relevance for web development
\item Static behavioural typing
\item Static channel linearity by construction
\item Pros - type dependencies (expressing dependent types)
\item Cons
\end{itemize}
\item Other initiatives -- creating language with first-class channel primitives (e.g. SILL)
\end{itemize}

PureScript - \cite{PureScript2019}

\subsection{Hybrid Session Verification}
\begin{itemize}
\item How the aforementioned 2 approaches motivated hybrid session verification
\item Workflow using Scribble toolchain 
\item Static behavioural typing, dynamic channel usage runtime checks
\item Pros - Abstract I/O interfaces
\item Pros - Input futures for `complex communication patterns' motivated by SMTP example
\item Cons - ``practical compromise'' as outlined in paper
\end{itemize}

Java - \cite{Hybrid2016}

>>>>

\subsection{Type-Level Encoding}
King et al. presented an approach in \cite{PureScript2019} for integrating session types into web development using the PureScript language, which takes advantage of its expressive type system to provide static guarantees on implementation conformance with respect to the protocol. We outline the main components of their EFSM encoding:

\paragraph{Actions as type classes} The semantics of the state transition function in the EFSM formalism express that a tuple of state-action \textit{uniquely defines} a successor state. These semantics can be expressed by \textit{multi-parameter type classes (MPTC) with functional dependencies}: \texttt{class Send r s t a | s -> t r a} defines \texttt{Send} as a MPTC parameterised by recipient \texttt{r}, current state \texttt{s}, successor state \texttt{t} and payload type \texttt{a}, and \texttt{s -> t r a} expresses the functional dependency that, for an instance of this type class, the current state uniquely determines the successor state, the recipient and payload type. These type classes are independent of the EFSM.

\paragraph{Transitions as instances of type classes} By encoding states as data types, valid EFSM transitions are encoded as \textit{instances} of the type classes. If \texttt{S1} is an output state, sending an \texttt{Int} to \texttt{Svr} with successor \texttt{S2}, we would encode \texttt{instance SendS1 :: Send Svr S1 S2 Int}. Because of the functional dependency, the developer cannot instantiate an invalid transition (e.g. \texttt{Send Svr S1 S3 Bool}, since \texttt{S1} uniquely determines the other type parameters.

This proposal is relevant to the problem we are tackling, and we appreciate that the intricacy of the library design in the communication combinators to conceal the channel resources is something we can build upon in our solution. We also observe the challenges of applying session types to the front-end environment, as shown by the careful choice made in \cite{PureScript2019} to use the \textit{Concur UI} framework which builds UI elements sequentially to model sequential sessions; not doing so would require binding channel resources into event listeners on UI elements, which makes linearity violation possible (e.g. by binding a send transition to a button click event, the user might click the button twice, thus reusing the channel).

\subsection{Hybrid Session Verification}
Whilst runtime monitors and type-level encoding achieve communication safety guarantees via dynamic and static approaches respectively, Hu and Yoshida implemented a workflow in \cite{Hybrid2016} that performs hybrid verification of communication protocol conformance for Java applications. We highlight the two main components below:

\paragraph{Static session typing} States are represented as classes; supported transitions on each state are represented as instance methods, parameterised by the role, label and payload involved in the message exchange. A send method takes the payload to send as a parameter, whilst a receive method is a blocking call that requires the caller to allocate a \texttt{Buf<T>} wrapper on the stack (where \texttt{T} is the expected payload type), then the receive method populates the payload into the wrapper and returns upon receiving from the channel. These instance methods return a new instance of the successor state class.

\paragraph{Runtime linear channel usage checks} Each state keeps track of its usage in a private boolean flag and throws an exception when the instance method is called twice, which signifies channel reuse and a linearity violation. Similarly, the \texttt{SessionEndpoint} class keeps track of whether the connection is open or close, and throws an exception when program execution exits the scope of the session endpoint and a terminal state has yet to be reached, signifying a protocol violation as the session hasn't completed yet but is now out of scope.

We find that this proposal strikes a good balance between maximising static communication safety guarantees whilst providing an intuitive set of APIs for developers to efficiently write their applications. For example, the encoding of transitions as instance methods that return new instances of the successor state exposes the channel resource, and since Java does not build support for linear resources and does not monitor aliasing of variables, linear channel usage is monitored dynamically. However, this complements the imperative style of application code written in Java, and takes advantage of the type system to statically enforce valid transitions in the code.

### 2020-01-23 interim/sections/background/scribble.tex
Interim: add introduction and lstlisting styles from PLACES

\subsection{Overview}
% Introduce Scribble as an implementation of MPST
Whilst session type theory represents the type language for concurrent processes, it also forms the theoretical basis of proposals introduced to implement session types for real-world application development: the Scribble language is one such proposal.

@@

\begin{figure}[!h]
\begin{lstlisting}
type <java> "java.lang.Integer" from "rt.jar" as int;

@@

% Formalise syntax and properties of EFSM derived from well-formed protocols
To facilitate the verification process, a local protocol is converted into 
an \term{endpoint finite state machine} (EFSM). An EFSM encodes the control flow of the local protocol, where an initial state is defined and each transition from some state to a successor state corresponds to a valid IO action (i.e. sending to or receiving from another role) permitted at the endpoint at that state.

@@

\end{array}
\]
\singlespacing
\caption{Syntax for Endpoint FSM}
\end{figure}

\begin{figure}[!hb]
\doublespacing
\[
\begin{array}{rl}

\texttt{initial}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
\texttt{terminal}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \emptyset \\
\texttt{output}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\} \\
\texttt{input}_{\mathbb{R}, \mathbb{L}, \mathbb{T}, \Sigma, \mathbb{S}, \delta}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists r \in \mathbb{R}, l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\} \\
\end{array}
\]
\singlespacing
\caption{Types of EFSM States}
\end{figure}

>>>>

%\subsection{Overview}
% Introduce Scribble as an implementation of MPST
Whilst session type theory represents the type language for concurrent processes, it also forms the theoretical basis of proposals introduced to implement session types for real-world application development: the Scribble language is one such proposal.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=Scribble]
type <java> "java.lang.Integer" from "rt.jar" as int;

@@

% Formalise syntax and properties of EFSM derived from well-formed protocols
To facilitate the verification process, a local protocol is converted into 
an \term{endpoint finite state machine} (EFSM). An EFSM encodes the control flow of the local protocol, where an initial state is defined and each transition from some state to a successor state corresponds to a valid IO action (i.e. sending to or receiving from another role) permitted at the endpoint at that state. Figure \ref{fig:efsmsyntax} presents the EFSM formalism adopted from \cite{Hybrid2016}.

@@

\end{array}
\]
\singlespacing
\captionof{figure}{Syntax for Endpoint FSM}
\label{fig:efsmsyntax}
\end{figure}

We highlight the semantics of the \term{state transition function}. Its domain defines the set of permitted actions for each state, and the range defines the successor state. Crucially, this is a \term{partial function} because the EFSM restricts what can be done (in terms of send and receive actions, and what label/payload types can be sent and received).

\begin{figure}[!h]
\doublespacing
\[
\begin{array}{rl}

\texttt{initial}_{EFSM}(S) \iff & \nexists S' \in \mathbb{S},\alpha \in \Sigma.~\delta(S',\alpha) = S \\
\texttt{terminal}_{EFSM}(S) \iff & \delta(S) = \emptyset \\
\texttt{output}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r!l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R} \\
\texttt{input}_{EFSM}(S) \iff & \delta(S) = \{\,\alpha \in \Sigma \mid \exists l \in \mathbb{L}, \tilde{T} \subseteq{\mathbb{T}}.~\alpha = r?l(\tilde{T})\,\}~\text{for some}~r \in \mathbb{R}
\end{array}
\]
\singlespacing
\captionof{figure}{Types of EFSM States}
\label{fig:efsmstates}
\end{figure}

\cite{Hybrid2016} also identifies properties of EFSMs for well-formed protocol specifications: \textbf{(1)} there is exactly one initial state, \textbf{(2)} there is at most one terminal state, and \textbf{(3)} every non-terminal state is either an input state or output state. There are further restrictions on input and output states, namely these states must only send to or receive from the same role. We formalise these properties in Figure \ref{fig:efsmstates}.
